---
title: "Replication in Gemstones models (final step)"
author: "Sean Trott"
date: "August 15, 2025"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "pdf")
```


```{r include=FALSE}
library(tidyverse)
library(lme4)
library(viridis)
library(ggridges)
library(lmerTest)
library(ggrepel)
library(ggcorrplot)
library(vroom)
library(ggeffects)
library(broom.mixed)
library(kableExtra)
library(MuMIn)

all_colors <- viridis::viridis(10, option = "mako")
my_colors <- all_colors[c(3, 5, 7)]  # Selecting specific colors from the palette
```

# Induction analysis

## Load Gemstones data


```{r include=FALSE}
## setwd("/Users/seantrott/Dropbox/UCSD/Research/NLMs/epistemology/mechinterp_generalizability/src/analysis")
directory_path <- "../../data/processed/induction_results_gemstones/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
length(csv_files)
csv_list <- csv_files %>%
  map(~ read_csv(.))


df_models <- bind_rows(csv_list) %>%
  mutate(model = sub("Gemstone-Models/Gemstone-", "", mpath))  %>%
  mutate(layer = layer + 1)
nrow(df_models)


### Create head-level summaries
df_gemstones_models_head_summaries = df_models %>%
  group_by(mpath, n_params, model,
           layer, head) %>%
  summarise(mean_ps = mean(ps),
            mean_ar_acc = mean(ar_acc),
            mean_ar_rank = mean(ar_mean_rank))

```




## Positional analysis

```{r induction_head_pos_gemstones}


### Max by layer
df_gemstones_models_head_summaries %>%
  group_by(model, head, layer, n_params) %>%
  summarise(max_mean = max(mean_ps)) %>%
  ggplot(aes(x = layer,
             y = head,
             fill = max_mean)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Head",
       fill = "Max Prefix-Matching Score") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "Max Prefix-Matching Score") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~reorder(model, n_params))




### How does *max* attention per layer change across model size?
df_layerwise_attn <- df_gemstones_models_head_summaries %>%
  group_by(mpath) %>%
  mutate(z_ps = scale(mean_ps)) %>%
  group_by(mpath, layer, n_params) %>%
  summarise(
    max_attention = max(z_ps),
    se_attention = sd(z_ps) / sqrt(n()),
    .groups = "drop"
  )

### 
summary(lm(data = df_layerwise_attn,
             max_attention ~ log10(n_params) * layer))



### Relative layer
df_gemstones_models_head_summaries = df_gemstones_models_head_summaries %>%
  group_by(model) %>%
  mutate(max_layer = max(layer),
         prop_layer = layer / max_layer) %>%
  ### Scale for interpreting the coefficients more easily
  mutate(prop_layer_scaled = scale(prop_layer)) %>%
  ungroup()






### Layer depth across all models
df_gemstones_models_head_summaries %>%
  group_by(model, prop_layer) %>%
  summarise(max_ps = max(mean_ps), .groups = "drop") %>%
  mutate(bin = cut(prop_layer, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
         bin_mid = as.numeric(sub("[\\(\\[]([0-9.]+).*", "\\1", bin)) + 0.05) %>%
  group_by(bin_mid) %>%
  summarise(mean_ps = mean(max_ps),
            se = sd(max_ps)/sqrt(n()),
            .groups = "drop") %>%
  ggplot(aes(x = bin_mid, y = mean_ps)) +
  geom_ribbon(aes(ymin = mean_ps - se, ymax = mean_ps + se), alpha = 0.3) +
  geom_line(size = 1.2) +
  labs(x = "Layer depth", y = "Max prefix-matching score") +
  theme_minimal(base_size = 15)



### Visualizing relative layer
summary_df <- df_gemstones_models_head_summaries %>%
  mutate(binned_prop_layer = ntile(prop_layer, 6)) %>%
  mutate(prop_binned = binned_prop_layer / 6) %>%
  group_by(model, n_params, prop_binned) %>%
  summarise(
    max_ratio_per_seed = max(mean_ps, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(model, prop_binned, n_params) %>%
  summarise(
    avg_max_ratio = mean(max_ratio_per_seed, na.rm = TRUE),
    se_max_ratio = sd(max_ratio_per_seed, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

### For each model
ggplot(summary_df, aes(x = prop_binned, y = avg_max_ratio, 
                       color = reorder(model, n_params))) +
  geom_line(size = .5) +
  labs(x = "Layer Depth",
       y = "Max Prefix-Matching Score",
       color = "", fill = "") +
  scale_color_manual(values = viridisLite::viridis(22, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  scale_fill_manual(values = viridisLite::viridis(22, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  theme_minimal(base_size = 15) +
  theme(text = element_text(size = 15),
        legend.position = "none") 



### Analyses
summary(lmer(mean_ps ~ prop_layer + (1|head) + (1|model), 
             data = df_gemstones_models_head_summaries))

```



### Relative vs. absolute position

```{r gemstones_peak_layer}
### Relative vs. absolute position

# Identify the layer with peak PS for each model
df_peak_layer = df_gemstones_models_head_summaries %>%
  mutate(
    width = as.numeric(str_extract(model, "(?<=Gemstone-)\\d+"))
  ) %>%
  group_by(model, mpath, n_params, width) %>%
  mutate(max_layer = max(layer)) %>%
  mutate(aspect_ratio = width / max_layer) %>%
  slice_max(mean_ps, n = 1, with_ties = FALSE) %>%
  mutate(peak_layer_depth = layer / max_layer) %>%
  select(model, mpath, n_params, max_layer, peak_layer = layer, 
         peak_layer_depth, aspect_ratio, width) %>%
  ungroup()


### Relative layer depth
df_peak_layer %>%
  ggplot(aes(x = peak_layer_depth)) +
  geom_density(fill = my_colors[2], alpha = 0.6, color = NA) +
  geom_vline(xintercept = median(df_peak_layer$peak_layer_depth),
             linetype = "dashed") +
  geom_rug(length = unit(0.05, "npc")) +
  labs(x = "Relative layer depth of peak prefix-matching score",
       y = "Density") +
  xlim(0, 1) +
  theme_minimal(base_size = 15)


### 


# How does the number of possible layers relate to 
# *peak relative layer* and *peak layer*?
m_relative = lm(peak_layer_depth ~ log10(max_layer), data = df_peak_layer)
m_absolute = lm(peak_layer ~ log10(max_layer), data = df_peak_layer)
summary(m_relative)  
summary(m_absolute) 



df_peak_layer %>%
  ggplot(aes(x = max_layer,
             y = peak_layer_depth)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "Number of Layers (Log10)",
       y = "Layer Depth of Peak PS") +
  theme_minimal(base_size = 15) +
  theme(text = element_text(size = 15),
        legend.position = "none") 


df_peak_layer %>%
  ggplot(aes(x = max_layer,
             y = peak_layer)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "Number of Layers",
       y = "Layer with Peak PS") +
  theme_minimal(base_size = 15) +
  theme(text = element_text(size = 15),
        legend.position = "none") 


```


# 1-back and first-token

## Load Gemstones data


```{r include=FALSE}
## setwd("/Users/seantrott/Dropbox/UCSD/Research/NLMs/epistemology/mechinterp_generalizability/src/analysis")
directory_path <- "../../data/processed/attention_gemstones_summaries/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
length(csv_files)
csv_list <- csv_files %>%
  map(~ read_csv(.))


df_gemstones_models_1back_summaries <- bind_rows(csv_list) %>%
  mutate(model = sub("Gemstone-Models/Gemstone-", "", mpath))  %>%
    filter(Layer > 0) %>%
  mutate(head = Head - 1,
         layer = Layer)
  
nrow(df_models)




```

## 1-back

```{r 1back_babylm_pos}

### Max by layer
df_gemstones_models_1back_summaries %>%
  group_by(model, head, layer, n_params) %>%
  summarise(max_mean = max(mean_1back)) %>%
  ggplot(aes(x = layer,
             y = head,
             fill = max_mean)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Head",
       fill = "Max 1-back Attention") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "Max 1-back Attention") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~reorder(model, n_params))



```


### Relative vs. absolute position

```{r gemstones_peak_layer_1back}
### Relative vs. absolute position

# Identify the layer with peak 1back for each model
df_peak_layer = df_gemstones_models_1back_summaries %>%
  mutate(
    width = as.numeric(str_extract(model, "(?<=Gemstone-)\\d+"))
  ) %>%
  group_by(model, n_params, width) %>%
  mutate(max_layer = max(layer)) %>%
  mutate(aspect_ratio = width / max_layer) %>%
  slice_max(mean_1back, n = 1, with_ties = FALSE) %>%
  mutate(peak_layer_depth = layer / max_layer) %>%
  select(model, n_params, max_layer, peak_layer = layer, 
         peak_layer_depth, aspect_ratio, width) %>%
  ungroup()


### Relative layer depth
df_peak_layer %>%
  ggplot(aes(x = peak_layer_depth)) +
  geom_density(fill = my_colors[2], alpha = 0.6, color = NA) +
  geom_vline(xintercept = median(df_peak_layer$peak_layer_depth),
             linetype = "dashed") +
  geom_rug(length = unit(0.05, "npc")) +
  labs(x = "Relative layer depth of peak 1-back strength",
       y = "Density") +
  xlim(0, 1) +
  theme_minimal(base_size = 15)


### 


# How does the number of possible layers relate to 
# *peak relative layer* and *peak layer*?
m_relative = lm(peak_layer_depth ~ log10(max_layer), data = df_peak_layer)
m_absolute = lm(peak_layer ~ log10(max_layer), data = df_peak_layer)
summary(m_relative)  
summary(m_absolute) 



df_peak_layer %>%
  ggplot(aes(x = max_layer,
             y = peak_layer_depth)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "Number of Layers (Log10)",
       y = "Layer Depth of Peak 1-back") +
  theme_minimal(base_size = 15) +
  theme(text = element_text(size = 15),
        legend.position = "none") 


df_peak_layer %>%
  ggplot(aes(x = max_layer,
             y = peak_layer)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "Number of Layers",
       y = "Layer with Peak PS") +
  theme_minimal(base_size = 15) +
  theme(text = element_text(size = 15),
        legend.position = "none") 


```


## First-token


### Relative vs. absolute position

```{r gemstones_peak_layer_first_token}
### Relative vs. absolute position

# Identify the layer with peak 1back for each model
df_peak_layer = df_gemstones_models_1back_summaries %>%
  mutate(
    width = as.numeric(str_extract(model, "(?<=Gemstone-)\\d+"))
  ) %>%
  group_by(model, n_params, width) %>%
  mutate(max_layer = max(layer)) %>%
  mutate(aspect_ratio = width / max_layer) %>%
  slice_max(mean_first, n = 1, with_ties = FALSE) %>%
  mutate(peak_layer_depth = layer / max_layer) %>%
  select(model, n_params, max_layer, peak_layer = layer, 
         peak_layer_depth, aspect_ratio, width) %>%
  ungroup()


### Relative layer depth
df_peak_layer %>%
  ggplot(aes(x = peak_layer_depth)) +
  geom_density(fill = my_colors[2], alpha = 0.6, color = NA) +
  geom_vline(xintercept = median(df_peak_layer$peak_layer_depth),
             linetype = "dashed") +
  geom_rug(length = unit(0.05, "npc")) +
  labs(x = "Relative layer depth of peak first token strength",
       y = "Density") +
  xlim(0, 1) +
  theme_minimal(base_size = 15)


### 


# How does the number of possible layers relate to 
# *peak relative layer* and *peak layer*?
m_relative = lm(peak_layer_depth ~ log10(max_layer), data = df_peak_layer)
m_absolute = lm(peak_layer ~ log10(max_layer), data = df_peak_layer)
summary(m_relative)  
summary(m_absolute) 



df_peak_layer %>%
  ggplot(aes(x = max_layer,
             y = peak_layer_depth)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "Number of Layers (Log10)",
       y = "Layer Depth of Peak First Token") +
  theme_minimal(base_size = 15) +
  theme(text = element_text(size = 15),
        legend.position = "none") 


df_peak_layer %>%
  ggplot(aes(x = max_layer,
             y = peak_layer)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "Number of Layers",
       y = "Layer with Peak First Token") +
  theme_minimal(base_size = 15) +
  theme(text = element_text(size = 15),
        legend.position = "none") 


```


# Plotting all together

## Max scores across layers

```{r gemstones_all_scores_layers_max}
df_1back_prop <- df_gemstones_models_1back_summaries %>%
  group_by(model) %>%
  mutate(max_layer = max(layer),
         prop_layer = layer / max_layer) %>%
  ungroup()

df_ps_max <- df_gemstones_models_head_summaries %>%
  group_by(model, prop_layer) %>%
  summarise(max_val = max(mean_ps), .groups = "drop") %>%
  mutate(bin = cut(prop_layer, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
         bin_mid = as.numeric(sub("[\\(\\[]([0-9.]+).*", "\\1", bin)) + 0.05) %>%
  group_by(bin_mid) %>%
  summarise(mean_val = mean(max_val), se = sd(max_val) / sqrt(n()), .groups = "drop") %>%
  mutate(metric = "Prefix-matching (PS)")

df_1back_max <- df_1back_prop %>%
  group_by(model, prop_layer) %>%
  summarise(max_val = max(mean_1back), .groups = "drop") %>%
  mutate(bin = cut(prop_layer, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
         bin_mid = as.numeric(sub("[\\(\\[]([0-9.]+).*", "\\1", bin)) + 0.05) %>%
  group_by(bin_mid) %>%
  summarise(mean_val = mean(max_val), se = sd(max_val) / sqrt(n()), .groups = "drop") %>%
  mutate(metric = "1-back")

df_first_max <- df_1back_prop %>%
  group_by(model, prop_layer) %>%
  summarise(max_val = max(mean_first), .groups = "drop") %>%
  mutate(bin = cut(prop_layer, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
         bin_mid = as.numeric(sub("[\\(\\[]([0-9.]+).*", "\\1", bin)) + 0.05) %>%
  group_by(bin_mid) %>%
  summarise(mean_val = mean(max_val), se = sd(max_val) / sqrt(n()), .groups = "drop") %>%
  mutate(metric = "First-token")

df_depth_all <- bind_rows(df_ps_max, df_1back_max, df_first_max) %>%
  mutate(metric = factor(metric,
                         levels = c("First-token", "1-back", "Prefix-matching (PS)")))

df_depth_all %>%
  ggplot(aes(x = bin_mid, y = mean_val, color = metric, fill = metric)) +
  geom_ribbon(aes(ymin = mean_val - se, ymax = mean_val + se),
              alpha = 0.3, color = NA) +
  geom_line(size = 1.2) +
  scale_color_manual(values = my_colors) +
  scale_fill_manual(values = my_colors) +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = "Layer depth", y = "Mean of per-model max", color = NULL, fill = NULL) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none")
```

## Mean score

```{r gemstones_all_scores_layers_mean}
df_ps_mean <- df_gemstones_models_head_summaries %>%
  group_by(model, prop_layer) %>%
  summarise(mean_per_model = mean(mean_ps), .groups = "drop") %>%
  mutate(bin = cut(prop_layer, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
         bin_mid = as.numeric(sub("[\\(\\[]([0-9.]+).*", "\\1", bin)) + 0.05) %>%
  group_by(bin_mid) %>%
  summarise(mean_val = mean(mean_per_model), se = sd(mean_per_model) / sqrt(n()), .groups = "drop") %>%
  mutate(metric = "Prefix-matching (PS)")

df_1back_mean <- df_1back_prop %>%
  group_by(model, prop_layer) %>%
  summarise(mean_per_model = mean(mean_1back), .groups = "drop") %>%
  mutate(bin = cut(prop_layer, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
         bin_mid = as.numeric(sub("[\\(\\[]([0-9.]+).*", "\\1", bin)) + 0.05) %>%
  group_by(bin_mid) %>%
  summarise(mean_val = mean(mean_per_model), se = sd(mean_per_model) / sqrt(n()), .groups = "drop") %>%
  mutate(metric = "1-back")

df_first_mean <- df_1back_prop %>%
  group_by(model, prop_layer) %>%
  summarise(mean_per_model = mean(mean_first), .groups = "drop") %>%
  mutate(bin = cut(prop_layer, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
         bin_mid = as.numeric(sub("[\\(\\[]([0-9.]+).*", "\\1", bin)) + 0.05) %>%
  group_by(bin_mid) %>%
  summarise(mean_val = mean(mean_per_model), se = sd(mean_per_model) / sqrt(n()), .groups = "drop") %>%
  mutate(metric = "First-token")

df_depth_all_mean <- bind_rows(df_ps_mean, df_1back_mean, df_first_mean) %>%
  mutate(metric = factor(metric,
                         levels = c("First-token", "1-back", "Prefix-matching (PS)")))

df_depth_all_mean %>%
  ggplot(aes(x = bin_mid, y = mean_val, color = metric, fill = metric)) +
  geom_ribbon(aes(ymin = mean_val - se, ymax = mean_val + se),
              alpha = 0.3, color = NA) +
  geom_line(size = 1.2) +
  scale_color_manual(values = my_colors) +
  scale_fill_manual(values = my_colors) +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = "Layer depth", y = "Mean attention across heads", color = NULL, fill = NULL) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none")
```


## Peak layer depth

```{r gemstones_all_peaks}

peak_ps <- df_gemstones_models_head_summaries %>%
  group_by(model, n_params, max_layer) %>%
  mutate(max_layer = max(layer)) %>%
  slice_max(mean_ps, n = 1, with_ties = FALSE) %>%
  transmute(model, n_params, max_layer,
            peak_layer_depth = layer / max_layer,
            metric = "Prefix-matching (PS)") %>%
  ungroup()

peak_1back <- df_gemstones_models_1back_summaries %>%
  group_by(model, n_params) %>%
  mutate(max_layer = max(layer)) %>%
  slice_max(mean_1back, n = 1, with_ties = FALSE) %>%
  transmute(model, n_params, max_layer,
            peak_layer_depth = layer / max_layer,
            metric = "1-back") %>%
  ungroup()

peak_first <- df_gemstones_models_1back_summaries %>%
  group_by(model, n_params) %>%
  mutate(max_layer = max(layer)) %>%
  slice_max(mean_first, n = 1, with_ties = FALSE) %>%
  transmute(model, n_params, max_layer,
            peak_layer_depth = layer / max_layer,
            metric = "First-token") %>%
  ungroup()

df_peak_all <- bind_rows(peak_ps, peak_1back, peak_first) %>%
  mutate(metric = factor(metric,
                         levels = c("First-token", "1-back", "Prefix-matching (PS)")))

df_peak_all %>%
  ggplot(aes(x = peak_layer_depth, y = metric, fill = metric)) +
  geom_density_ridges(alpha = 0.7, color = NA, scale = 1.1,
                      rel_min_height = 0.01,
                      jittered_points = TRUE,
                      position = position_raincloud(height = 0, ygap = 0.05),
                      point_shape = "|", point_size = 4,
                      point_alpha = 1, point_color = "grey20") +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_fill_manual(values = my_colors) +
  xlim(0, 1) +
  labs(x = "Relative layer depth of peak", y = NULL) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none")


mod_full = lmer(data = df_peak_all %>% filter(metric != "First-token"),
                peak_layer_depth ~ log10(max_layer) + metric + (1|model))
summary(mod_full)
```


# Merge with other model families

## Load Pythia seeds, Pythia-main, and BabyLM (Final step)

```{r}
### Pythia seeds
pyt_seed_ps_dir <- "../../data/processed/induction_results/"
pyt_seed_1b_dir <- "../../data/processed/attention_summaries/"

pyt_seed_ps_files <- list.files(pyt_seed_ps_dir, pattern = "step143000-seed", full.names = TRUE)
pyt_seed_1b_files <- list.files(pyt_seed_1b_dir, pattern = "step143000-seed", full.names = TRUE)

length(pyt_seed_ps_files)
length(pyt_seed_1b_files)

df_pythia_seed_head_summaries <- pyt_seed_ps_files %>%
  map(read_csv) %>% bind_rows() %>%
  mutate(model = sub("^EleutherAI/", "", mpath), layer = layer + 1,
         seed_name = paste0("seed", seed)) %>%
  group_by(mpath, n_params, model, seed, seed_name, layer, head) %>%
  summarise(mean_ps = mean(ps), .groups = "drop")

df_pythia_seed_1back_summaries <- pyt_seed_1b_files %>%
  map(read_csv) %>% bind_rows() %>%
  mutate(model = sub("^EleutherAI/", "", mpath), head = Head - 1, layer = Layer,
         seed_name = paste0("seed", seed)) %>%
  filter(Layer > 0) %>%
  group_by(mpath, model, n_params, seed, seed_name, layer, head) %>%
  summarise(mean_1back = mean(mean_1back), mean_first = mean(mean_first), .groups = "drop")

table(df_pythia_seed_head_summaries$mpath, df_pythia_seed_head_summaries$seed_name)

### Pythia maijn
pyt_ps_dir   <- "../../data/processed/induction_results_main/"
pyt_1b_dir   <- "../../data/processed/attention_main_summaries/"

pyt_ps_files <- list.files(pyt_ps_dir, pattern = "step143000\\.csv$", full.names = TRUE)
pyt_1b_files <- list.files(pyt_1b_dir, pattern = "step143000.*\\.csv$", full.names = TRUE)

df_pythia_models_head_summaries <- pyt_ps_files %>%
  map(read_csv) %>% bind_rows() %>%
  mutate(model = sub("^EleutherAI/", "", mpath), layer = layer + 1) %>%
  group_by(mpath, n_params, model, layer, head) %>%
  summarise(mean_ps = mean(ps), .groups = "drop")

df_pythia_models_1back_summaries <- pyt_1b_files %>%
  map(read_csv) %>% bind_rows() %>%
  mutate(model = sub("^EleutherAI/", "", mpath), head = Head - 1, layer = Layer) %>%
  filter(Layer > 0) %>%
  group_by(mpath, model, n_params, layer, head) %>%
  summarise(mean_1back = mean(mean_1back), mean_first = mean(mean_first), .groups = "drop")

baby_ps_dir <- "../../data/processed/induction_results_babylm/"
baby_1b_dir <- "../../data/processed/attentions_babylm/"

baby_ps_files <- list.files(baby_ps_dir, pattern = "-main\\.csv$", full.names = TRUE)
baby_1b_files <- list.files(baby_1b_dir, pattern = "-main\\.csv$", full.names = TRUE)

df_babylm_models_head_summaries <- baby_ps_files %>%
  map(read_csv) %>% bind_rows() %>%
  mutate(model = sub("BabyLM-community/", "", mpath), layer = layer + 1) %>%
  group_by(mpath, n_params, model, layer, head) %>%
  summarise(mean_ps = mean(ps), .groups = "drop")

df_babylm_models_1back_summaries <- baby_1b_files %>%
  map(read_csv) %>% bind_rows() %>%
  mutate(model = sub("BabyLM-community/", "", mpath), head = Head - 1, layer = Layer) %>%
  filter(Layer > 0) %>%
  group_by(mpath, model, n_params, layer, head) %>%
  summarise(mean_1back = mean(`1-back attention`), mean_first = mean(first_token), .groups = "drop")
```

## Merge

```{r}
build_peaks <- function(head_df, oneback_df, suite_name) {
  ps_long <- head_df %>%
    transmute(suite = suite_name, model, n_params, layer,
              metric = "Prefix-matching (PS)", value = mean_ps)

  ob_long <- oneback_df %>%
    transmute(suite = suite_name, model, n_params, layer,
              metric = "1-back", value = mean_1back)

  ft_long <- oneback_df %>%
    transmute(suite = suite_name, model, n_params, layer,
              metric = "First-token", value = mean_first)

  bind_rows(ps_long, ob_long, ft_long) %>%
    group_by(suite, model, n_params, metric) %>%
    mutate(max_layer = max(layer)) %>%
    slice_max(value, n = 1, with_ties = FALSE) %>%
    mutate(peak_layer_depth = layer / max_layer) %>%
    ungroup() %>%
    select(suite, model, n_params, max_layer, peak_layer = layer,
           peak_layer_depth, metric)
}

build_peaks_seeded <- function(head_df, oneback_df, suite_name) {
  ps_long <- head_df %>%
    transmute(suite = suite_name, model, seed_name, n_params, layer,
              metric = "Prefix-matching (PS)", value = mean_ps)
  ob_long <- oneback_df %>%
    transmute(suite = suite_name, model, seed_name, n_params, layer,
              metric = "1-back", value = mean_1back)
  ft_long <- oneback_df %>%
    transmute(suite = suite_name, model, seed_name, n_params, layer,
              metric = "First-token", value = mean_first)

  bind_rows(ps_long, ob_long, ft_long) %>%
    group_by(suite, model, seed_name, n_params, metric) %>%
    mutate(max_layer = max(layer)) %>%
    slice_max(value, n = 1, with_ties = FALSE) %>%
    mutate(peak_layer_depth = layer / max_layer) %>%
    ungroup() %>%
    select(suite, model, seed_name, n_params, max_layer,
           peak_layer = layer, peak_layer_depth, metric)
}

df_peak_seeds <- build_peaks_seeded(
  df_pythia_seed_head_summaries,
  df_pythia_seed_1back_summaries, "Pythia"
) %>%
  mutate(metric = factor(metric,
                         levels = c("First-token", "1-back", "Prefix-matching (PS)")))



df_peak_all <- bind_rows(
  build_peaks(df_gemstones_models_head_summaries,
              df_gemstones_models_1back_summaries, "Gemstones"),
  build_peaks(df_pythia_models_head_summaries,
              df_pythia_models_1back_summaries, "Pythia"),
  build_peaks(df_babylm_models_head_summaries,
              df_babylm_models_1back_summaries, "BabyLM")
) %>%
  mutate(metric = factor(metric,
                         levels = c("First-token", "1-back", "Prefix-matching (PS)")),
         suite = factor(suite, levels = c("BabyLM", "Pythia", "Gemstones")))


df_peak_all_seeded <- df_peak_all %>%
  mutate(seed_name = "seed_main") %>%
  bind_rows(df_peak_seeds) %>%
  mutate(metric = factor(metric,
                         levels = c("First-token", "1-back", "Prefix-matching (PS)")),
         suite = factor(suite))


df_peak_all_seeded = df_peak_all_seeded %>%
  mutate(m_seed = paste(model, seed_name))

```

## Analyze

```{r all_suites_position}

df_mod_seeded <- df_peak_all_seeded %>%
  filter(metric != "First-token")

mod_seeded <- lmer(peak_layer_depth ~ log10(max_layer) + metric +
                     (1 | suite) + (1 | model),
                   data = df_mod_seeded)
summary(mod_seeded)

r.squaredGLMM(mod_seeded)


df_peak_all_seeded %>%
  ggplot(aes(x = peak_layer_depth, y = metric, fill = metric)) +
  geom_density_ridges(alpha = 0.6, color = NA, scale = 1.1,
                      rel_min_height = 0.01,
                      jittered_points = TRUE,
                      position = position_points_jitter(width = 0, height = 0),
                      point_shape = 21, point_size = 2,
                      point_alpha = 1, aes(point_fill = suite),
                      point_color = "grey20") +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_fill_manual(values = my_colors) +
  scale_discrete_manual("point_fill", values = c("white", "grey60", "black")) +
  xlim(0, 1) +
  labs(x = "Relative layer depth of peak", y = NULL, point_fill = "Suite") +
  theme_minimal(base_size = 15) +
  theme(legend.position = "bottom") +
  guides(fill = "none")


```

## Plotting by layer

```{r all_suites_by_layer}
make_long <- function(head_df, oneback_df, suite_name, seeded = FALSE) {
  if (seeded) {
    ps <- head_df %>% transmute(suite = suite_name, model, seed_name, n_params,
                                layer, head, metric = "Prefix-matching (PS)", value = mean_ps)
    ob <- oneback_df %>% transmute(suite = suite_name, model, seed_name, n_params,
                                   layer, head, metric = "1-back", value = mean_1back)
  } else {
    ps <- head_df %>% transmute(suite = suite_name, model, seed_name = "seed_main", n_params,
                                layer, head, metric = "Prefix-matching (PS)", value = mean_ps)
    ob <- oneback_df %>% transmute(suite = suite_name, model, seed_name = "seed_main", n_params,
                                   layer, head, metric = "1-back", value = mean_1back)
  }
  bind_rows(ps, ob)
}

df_long_all <- bind_rows(
  make_long(df_gemstones_models_head_summaries, df_gemstones_models_1back_summaries, "Gemstones"),
  make_long(df_pythia_models_head_summaries,    df_pythia_models_1back_summaries,    "Pythia"),
  make_long(df_babylm_models_head_summaries,    df_babylm_models_1back_summaries,    "BabyLM"),
  make_long(df_pythia_seed_head_summaries,      df_pythia_seed_1back_summaries,      "Pythia", seeded = TRUE)
) %>%
  group_by(suite, model, seed_name, metric) %>%
  mutate(max_layer = max(layer), prop_layer = layer / max_layer) %>%
  ungroup()

df_long_all %>%
  group_by(suite, model, seed_name, metric, prop_layer) %>%
  summarise(val = max(value), .groups = "drop") %>%
  mutate(bin = cut(prop_layer, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
         bin_mid = as.numeric(sub("[\\(\\[]([0-9.]+).*", "\\1", bin)) + 0.05) %>%
  group_by(metric, suite, bin_mid) %>%
  summarise(mean_val = mean(val), se = sd(val) / sqrt(n()), .groups = "drop") %>%
  ggplot(aes(x = bin_mid, y = mean_val, color = metric, fill = metric)) +
  geom_ribbon(aes(ymin = mean_val - se, ymax = mean_val + se), alpha = 0.3, color = NA) +
  geom_line(size = 1.2) +
  scale_color_manual(values = my_colors) +
  scale_fill_manual(values = my_colors) +
  facet_wrap(~metric + suite, scales = "free_y") +
  labs(x = "Layer depth", y = "Mean of per-network max", color = NULL, fill = NULL) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none")


df_long_all %>%
  group_by(suite, model, seed_name, metric, prop_layer) %>%
  summarise(val = mean(value), .groups = "drop") %>%
  mutate(bin = cut(prop_layer, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
         bin_mid = as.numeric(sub("[\\(\\[]([0-9.]+).*", "\\1", bin)) + 0.05) %>%
  group_by(metric, suite, bin_mid) %>%
  summarise(mean_val = mean(val), se = sd(val) / sqrt(n()), .groups = "drop") %>%
  ggplot(aes(x = bin_mid, y = mean_val, color = metric, fill = metric)) +
  geom_ribbon(aes(ymin = mean_val - se, ymax = mean_val + se), alpha = 0.3, color = NA) +
  geom_line(size = 1.2) +
  scale_color_manual(values = my_colors) +
  scale_fill_manual(values = my_colors) +
  facet_wrap(~metric + suite, scales = "free_y") +
  labs(x = "Layer depth", y = "Mean attention across heads", color = NULL, fill = NULL) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none")



df_long_all %>%
  group_by(suite, model, seed_name, metric, prop_layer) %>%
  summarise(val = max(value), .groups = "drop") %>%
  mutate(bin = cut(prop_layer, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
         bin_mid = as.numeric(sub("[\\(\\[]([0-9.]+).*", "\\1", bin)) + 0.05) %>%
  group_by(metric, bin_mid) %>%
  summarise(mean_val = mean(val), se = sd(val) / sqrt(n()), .groups = "drop") %>%
  ggplot(aes(x = bin_mid, y = mean_val, color = metric, fill = metric)) +
  geom_ribbon(aes(ymin = mean_val - se, ymax = mean_val + se), alpha = 0.3, color = NA) +
  geom_line(size = 1.2) +
  scale_color_manual(values = my_colors) +
  scale_fill_manual(values = my_colors) +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = "Layer depth", y = "Mean of per-network max", color = NULL, fill = NULL) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none")


df_long_all %>%
  group_by(suite, model, seed_name, metric, prop_layer) %>%
  summarise(val = mean(value), .groups = "drop") %>%
  mutate(bin = cut(prop_layer, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
         bin_mid = as.numeric(sub("[\\(\\[]([0-9.]+).*", "\\1", bin)) + 0.05) %>%
  group_by(metric, bin_mid) %>%
  summarise(mean_val = mean(val), se = sd(val) / sqrt(n()), .groups = "drop") %>%
  ggplot(aes(x = bin_mid, y = mean_val, color = metric, fill = metric)) +
  geom_ribbon(aes(ymin = mean_val - se, ymax = mean_val + se), alpha = 0.3, color = NA) +
  geom_line(size = 1.2) +
  scale_color_manual(values = my_colors) +
  scale_fill_manual(values = my_colors) +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = "Layer depth", y = "Mean attention across heads", color = NULL, fill = NULL) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none")

df_long_all %>%
  group_by(suite, model, seed_name, metric, prop_layer) %>%
  summarise(val = max(value), .groups = "drop") %>%
  mutate(bin = cut(prop_layer, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
         bin_mid = as.numeric(sub("[\\(\\[]([0-9.]+).*", "\\1", bin)) + 0.05) %>%
  group_by(metric, suite, bin_mid) %>%
  summarise(mean_val = mean(val), se = sd(val) / sqrt(n()), .groups = "drop") %>%
  ggplot(aes(x = bin_mid, y = mean_val, color = metric, fill = metric)) +
  geom_ribbon(aes(ymin = mean_val - se, ymax = mean_val + se), alpha = 0.2, color = NA) +
  geom_line(size = 1.2) +
  scale_color_manual(values = my_colors) +
  scale_fill_manual(values = my_colors) +
  labs(x = "Layer depth", y = "Mean of per-network max", color = NULL, fill = NULL) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "bottom") +
  facet_wrap(~suite)

df_long_all %>%
  group_by(suite, model, seed_name, metric, prop_layer) %>%
  summarise(val = mean(value), .groups = "drop") %>%
  mutate(bin = cut(prop_layer, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
         bin_mid = as.numeric(sub("[\\(\\[]([0-9.]+).*", "\\1", bin)) + 0.05) %>%
  group_by(metric, bin_mid, suite) %>%
  summarise(mean_val = mean(val), se = sd(val) / sqrt(n()), .groups = "drop") %>%
  ggplot(aes(x = bin_mid, y = mean_val, color = metric, fill = metric)) +
  geom_ribbon(aes(ymin = mean_val - se, ymax = mean_val + se), alpha = 0.2, color = NA) +
  geom_line(size = 1.2) +
  scale_color_manual(values = my_colors) +
  scale_fill_manual(values = my_colors) +
  labs(x = "Layer depth", y = "Mean attention across heads", color = NULL, fill = NULL) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "bottom") +
  facet_wrap(~suite)




```

## Key analyses

Here, we do the most principled analysis, in which we ask whether across the entire dataset.

```{r large_model}

### Just to confirm, using raw values
mod_full_mean = lmer(data = df_long_all,
                     value ~ metric * prop_layer + log10(max_layer) * prop_layer + 
                       metric * log10(max_layer) +
                       (1|suite) + (1|model))
summary(mod_full_mean)

r.squaredGLMM(mod_full_mean)


### Coefficient plot
coef_df <- tidy(mod_full_mean, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(term = recode(term,
    "metricPrefix-matching (PS)"                             = "PS (vs 1-back)",
    "prop_layer"                                             = "Relative depth",
    "log10(max_layer)"                                       = "log10(layer count)",
    "metricPrefix-matching (PS):prop_layer"                  = "PS × Relative depth",
    "metricPrefix-matching (PS):log10(max_layer)"            = "PS × log10(layers)",
    "prop_layer:log10(max_layer)"                            = "Relative depth × log10(layers)"
  )) %>%
  mutate(term = fct_reorder(term, estimate))

coef_df %>%
  ggplot(aes(x = estimate, y = term, color = p.value < 0.01)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), size = 0.7) +
  scale_color_manual(values = c("FALSE" = my_colors[1], "TRUE" = my_colors[3]),
                     guide = "none") +
  labs(x = "Coefficient estimate (95% CI)", y = NULL) +
  theme_minimal(base_size = 14)


### Table


tbl_df <- tidy(mod_full_mean, effects = "fixed", conf.int = TRUE) %>%
  mutate(
    term = recode(term,
      "(Intercept)"                                            = "Intercept",
      "metricPrefix-matching (PS)"                             = "PS (vs 1-back)",
      "prop_layer"                                             = "Relative depth",
      "log10(max_layer)"                                       = "$\\log_{10}$(layer count)",
      "metricPrefix-matching (PS):prop_layer"                  = "PS $\\times$ Rel.\\ depth",
      "metricPrefix-matching (PS):log10(max_layer)"            = "PS $\\times \\log_{10}$(layers)",
      "prop_layer:log10(max_layer)"                            = "Rel.\\ depth $\\times \\log_{10}$(layers)"
    ),
    Estimate = sprintf("%.3f", estimate),
    SE       = sprintf("%.3f", std.error),
    `95\\% CI` = sprintf("[%.3f, %.3f]", conf.low, conf.high),
    p = case_when(
      p.value < 0.001 ~ "$< 0.001$",
      TRUE            ~ sprintf("%.3f", p.value)
    )
  ) %>%
  select(Term = term, Estimate, SE, `95\\% CI`, p)

kbl(tbl_df, format = "latex", booktabs = TRUE, escape = FALSE,
    align = "lrrrr",
    caption = "Fixed effects from the mixed-effects model predicting head-level attention from metric (1-back vs.\\ PS), relative layer depth, and $\\log_{10}$(layer count), with random intercepts for suite, model, and head-within-model.",
    label = "tab:mod_full_mean") %>%
  kable_styling(latex_options = c("booktabs", "scale_down")) %>%
  row_spec(0, bold = TRUE)
```


