---
title: "Analysis of QK-composition"
author: "Sean Trott"
date: "August 15, 2025"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "pdf")
```


```{r include=FALSE}
library(tidyverse)
library(lme4)
library(viridis)
library(ggridges)
library(lmerTest)
library(ggrepel)
library(ggcorrplot)
library(vroom)
library(ggeffects)
library(dplyr)
library(pROC)

all_colors <- viridis::viridis(10, option = "mako")
my_colors <- all_colors[c(3, 5, 7)]  # Selecting specific colors from the palette
```

# Load Pythia data

## Load QK scores heads

```{r include=FALSE}
## setwd("/Users/seantrott/Dropbox/UCSD/Research/NLMs/epistemology/mechinterp_generalizability/src/analysis")
directory_path <- "../../data/processed/composed_qk_frob_results/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
length(csv_files)
csv_list <- csv_files %>%
  map(~ read_csv(.))


df_pythia_models <- bind_rows(csv_list) %>%
  mutate(model = sub("^EleutherAI/", "", mpath))  %>%
  mutate(step_modded = step + 1) %>%
  mutate(layer = layer + 1) %>%
  mutate(partner_layer = partner_layer + 1) %>%
  mutate(seed_name = paste0("seed", seed))
nrow(df_pythia_models)


table(df_pythia_models$seed)
table(df_pythia_models$mpath)

```

## Load prefix-matching scores

```{r include=FALSE}
directory_path <- "../../data/processed/induction_results/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
length(csv_files)
csv_list <- csv_files %>%
  map(~ read_csv(.))


df_pythia_models_ps <- bind_rows(csv_list) %>%
  mutate(model = sub("^EleutherAI/", "", mpath))  %>%
  mutate(step_modded = step + 1) %>%
  mutate(layer = layer + 1)
nrow(df_pythia_models_ps)


table(df_pythia_models_ps$seed)
table(df_pythia_models_ps$mpath)



### Create head-level summaries
df_pythia_models_head_summaries_ps = df_pythia_models_ps %>%
  group_by(mpath, n_params, step, step_modded, seed, model,
           layer, head) %>%
  summarise(mean_ps = mean(ps),
            mean_ar_acc = mean(ar_acc),
            mean_ar_rank = mean(ar_mean_rank)) %>%
  mutate(seed_name = paste0("seed", seed))


```


# Analysis

## Final step

```{r qk_scores_final_step}
df_pythia_models_final = df_pythia_models %>%
  filter(step == 143000)

df_pythia_models_final %>%
  filter(model == "pythia-14m") %>%
  ggplot(aes(x = layer,
             y = head,
             fill = composed_qk_score)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Head",
       fill = "Composed QK Score") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "Composed QK Score") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~reorder(seed_name, n_params))




### Max by layer
df_pythia_models_final %>%
  group_by(mpath, seed, layer, n_params) %>%
  summarise(max_qk = max(composed_qk_score)) %>%
  ggplot(aes(x = layer,
             y = seed,
             fill = max_qk)) +
  geom_tile() +
  labs(x = "Layer",
       y = "Seed",
       fill = "Max Composed QK Score") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       midpoint = 0, 
                       name = "Max Composed QK Score") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        strip.text.y = element_text(angle = 0), # Keep facet labels horizontal
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~reorder(mpath, n_params))

```


## Over training

```{r qk_scores_training_dynamics}

### Track max QK at each time point for each seed, across layer/head
df_max_qk_by_step = df_pythia_models %>%
  filter(model == "pythia-14m") %>%
  group_by(model, n_params, step_modded, seed, seed_name) %>%
  slice_max(composed_qk_score)


summary_avg = df_max_qk_by_step %>%
  group_by(model, n_params, step_modded) %>%
  summarise(
    mean_across_seeds = mean(composed_qk_score)
  )

grand_avg <- df_max_qk_by_step %>%
  group_by(step_modded) %>%
  summarise(grand_mean = mean(composed_qk_score))

df_max_qk_by_step %>%
  ggplot(aes(x = step_modded,
             y = composed_qk_score,
             color = factor(seed_name))) +
  geom_line(size = .6, alpha = .4) +   # individual curves
  geom_line(data = summary_avg, 
            aes(x = step_modded, y = mean_across_seeds), 
            color = "black", size = 2) +      # per-model average
  # geom_line(data = grand_avg, aes(x = step_modded, y = grand_mean),
  #          color = "red", size = 1.2, inherit.aes = FALSE) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Max Composed QK Score",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(9, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  facet_wrap(~reorder(model, n_params))

```


## Frob score over training

```{r qk_scores_training_dynamics_frob}

### Track max QK at each time point for each seed, across layer/head
df_max_qk_by_step = df_pythia_models %>%
  filter(model == "pythia-14m") %>%
  group_by(model, n_params, step_modded, seed, seed_name) %>%
  slice_max(composed_qk_frob)


summary_avg = df_max_qk_by_step %>%
  group_by(model, n_params, step_modded) %>%
  summarise(
    mean_across_seeds = mean(composed_qk_frob)
  )

df_max_qk_by_step %>%
  ggplot(aes(x = step_modded,
             y = composed_qk_frob,
             color = factor(seed_name))) +
  geom_line(size = .6, alpha = .4) +   # individual curves
  geom_line(data = summary_avg, 
            aes(x = step_modded, y = mean_across_seeds), 
            color = "black", size = 2) +      # per-model average
  # geom_line(data = grand_avg, aes(x = step_modded, y = grand_mean),
  #          color = "red", size = 1.2, inherit.aes = FALSE) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Max Composed QK Score",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(9, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  facet_wrap(~reorder(model, n_params))

```

## Seeds of specializaton?

Is there any systematic difference between heads that eventually look like induction heads (based on threshold) and those that don't, *pre-transition*? It seems the answer is no.

```{r seeds_specialization}
# Mark which heads become induction heads at the final checkpoint
INDUCTION_THRESHOLD <- 0.8

induction_status <- df_pythia_models %>%
  filter(step == 143000, model == "pythia-14m") %>%
  mutate(is_induction_final = composed_qk_score > INDUCTION_THRESHOLD) %>%
  select(seed_name, layer, head, is_induction_final)

# Annotate all checkpoints with future status
df_with_future <- df_pythia_models %>%
  filter(model == "pythia-14m") %>%
  inner_join(induction_status, by = c("seed_name", "layer", "head"))

# Plot: composed-QK over training, split by future induction status
df_with_future %>%
  group_by(step_modded, is_induction_final) %>%
  summarise(mean_qk = mean(composed_qk_score),
            se_qk = sd(composed_qk_score) / sqrt(n()),
            .groups = "drop") %>%
  ggplot(aes(x = step_modded, y = mean_qk, color = is_induction_final)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = mean_qk - se_qk, ymax = mean_qk + se_qk,
                  fill = is_induction_final), alpha = 0.2, color = NA) +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted") +
  labs(x = "Training Step (Log10)", y = "Mean Composed QK Score",
       color = "Future induction head", fill = "Future induction head") +
  theme_minimal() +
  theme(text = element_text(size = 14), legend.position = "bottom")


## Is there any signal in early steps?
very_early <- df_with_future %>% filter(step_modded <= 256)

mod_very_early <- lmer(composed_qk_score ~ is_induction_final +
                       (1 | seed_name) + (1 | step_modded),
                       data = very_early)
summary(mod_very_early)
```


# Correlation with prefix-matching scores

```{r}
### Merge data
df_merged_qk_and_ps = df_pythia_models_head_summaries_ps %>%
  inner_join(df_pythia_models)
nrow(df_merged_qk_and_ps)


```

## Final step

```{r merged_final_step}
df_merged_qk_and_ps_final = df_merged_qk_and_ps %>%
  filter(step == 143000)

df_merged_qk_and_ps_final %>%
  filter(model == "pythia-14m") %>%
  ggplot(aes(x = composed_qk_score,
             y = mean_ps,
             color = factor(layer))) +
  geom_point(alpha = .6) +
  geom_smooth(color = "gray") +
  labs(x = "Composed QK Score",
       y = "Mean Prefix-Matching Score",
       color = "Layer") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(5, option = "magma", 
                                                   begin = 0.8, end = 0.15)) + 
  facet_wrap(~reorder(model, n_params))

df_merged_qk_and_ps_final %>%
  filter(model == "pythia-14m") %>%
  ggplot(aes(x = composed_qk_score,
             y = mean_ps,
             color = factor(layer))) +
  geom_point(alpha = .6) +
  geom_smooth(color = "gray") +
  labs(x = "Composed QK Score",
       y = "Mean Prefix-Matching Score",
       color = "Layer") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(5, option = "magma", 
                                                   begin = 0.8, end = 0.15)) + 
  facet_wrap(~reorder(seed_name, n_params))


mod_full = lmer(data = df_merged_qk_and_ps_final,
             mean_ps ~ composed_qk_score + layer + (1|seed_name),
             REML = FALSE)

mod_reduced = lmer(data = df_merged_qk_and_ps_final,
             mean_ps ~ layer + (1|seed_name),
             REML = FALSE)

summary(mod_full)
anova(mod_full,mod_reduced)



df_merged_qk_and_ps_final %>%
  group_by(model, n_params, seed_name) %>%
  summarise(r_spearman = cor(mean_ps, composed_qk_score, method = "spearman"),
            r_pearson = cor(mean_ps, composed_qk_score, method = "pearson"))

df_merged_qk_and_ps_final %>%
  group_by(model, n_params, seed_name) %>%
  summarise(r_spearman = cor(mean_ps, composed_qk_score, method = "spearman"),
            r_pearson = cor(mean_ps, composed_qk_score, method = "pearson")) %>%
  summarise(mean(r_spearman),
            mean(r_pearson))


```

## Final step (frob)

```{r merged_final_step_frob}
df_merged_qk_and_ps_final = df_merged_qk_and_ps %>%
  filter(step == 143000)

df_merged_qk_and_ps_final %>%
  filter(model == "pythia-14m") %>%
  ggplot(aes(x = composed_qk_frob,
             y = mean_ps,
             color = factor(layer))) +
  geom_point(alpha = .6) +
  geom_smooth(color = "gray") +
  labs(x = "Composed QK Frob",
       y = "Mean Prefix-Matching Score",
       color = "Layer") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(5, option = "magma", 
                                                   begin = 0.8, end = 0.15)) + 
  facet_wrap(~reorder(model, n_params))

df_merged_qk_and_ps_final %>%
  filter(model == "pythia-14m") %>%
  ggplot(aes(x = composed_qk_frob,
             y = mean_ps,
             color = factor(layer))) +
  geom_point(alpha = .6) +
  geom_smooth(color = "gray") +
  labs(x = "Composed QK Frob",
       y = "Mean Prefix-Matching Score",
       color = "Layer") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(5, option = "magma", 
                                                   begin = 0.8, end = 0.15)) + 
  facet_wrap(~reorder(seed_name, n_params))


mod_full = lmer(data = df_merged_qk_and_ps_final,
             mean_ps ~ composed_qk_frob + layer + (1|seed_name),
             REML = FALSE)

mod_reduced = lmer(data = df_merged_qk_and_ps_final,
             mean_ps ~ layer + (1|seed_name),
             REML = FALSE)

summary(mod_full)
anova(mod_full,mod_reduced)



df_merged_qk_and_ps_final %>%
  group_by(model, n_params, seed_name) %>%
  summarise(r_spearman = cor(mean_ps, composed_qk_frob, method = "spearman"),
            r_pearson = cor(mean_ps, composed_qk_frob, method = "pearson"))

df_merged_qk_and_ps_final %>%
  group_by(model, n_params, seed_name) %>%
  summarise(r_spearman = cor(mean_ps, composed_qk_frob, method = "spearman"),
            r_pearson = cor(mean_ps, composed_qk_frob, method = "pearson")) %>%
  summarise(mean(r_spearman),
            mean(r_pearson))


cor.test(df_merged_qk_and_ps_final$composed_qk_frob,
         df_merged_qk_and_ps_final$mean_ps)

cor.test(df_merged_qk_and_ps_final$composed_qk_frob,
         df_merged_qk_and_ps_final$mean_ps, method = "spearman")


```


## Classify heads using metrics

Here, we label heads as induction or not and ask whether the two metrics combined predict their status.

```{r}
### First, regression (Even though not ideal)
summary(lm(data = df_merged_qk_and_ps_final,
           mean_ps ~ composed_qk_frob + composed_qk_score))$adj.r.squared
summary(lm(data = df_merged_qk_and_ps_final,
           mean_ps ~ composed_qk_frob ))$adj.r.squared
summary(lm(data = df_merged_qk_and_ps_final,
           mean_ps ~ composed_qk_score ))$adj.r.squared



### Now use classifier
ps_thresh <- 0.8

model_df <- df_merged_qk_and_ps_final %>%
  ungroup() %>%
  filter(model == "pythia-14m", step == 143000) %>%
  transmute(seed, layer, head,
            qk_eig  = composed_qk_score,
            qk_frob = composed_qk_frob,
            is_induction = as.integer(mean_ps >= ps_thresh))

table(model_df$is_induction)
table(model_df$is_induction) / nrow(model_df)


# --- leave-one-seed-out CV AUC for a given formula ---
seeds <- sort(unique(model_df$seed))
cv_auc <- function(formula) {
  preds <- lapply(seeds, function(s) {
    tr <- filter(model_df, seed != s); te <- filter(model_df, seed == s)
    fit <- glm(formula, data = tr, family = binomial)
    te$prob <- predict(fit, newdata = te, type = "response")
    te
  }) %>% bind_rows()
  as.numeric(auc(roc(preds$is_induction, preds$prob, quiet = TRUE)))
}

cv_auc(is_induction ~ qk_eig)              # eigenvalue alone
cv_auc(is_induction ~ qk_frob)             # Frobenius alone
cv_auc(is_induction ~ qk_eig + qk_frob)    # both



### Precision/recall
cv_preds <- function(formula) {
  lapply(seeds, function(s) {
    tr <- filter(model_df, seed != s); te <- filter(model_df, seed == s)
    fit <- glm(formula, data = tr, family = binomial)
    te$prob <- predict(fit, newdata = te, type = "response")
    te
  }) %>% bind_rows()
}

p <- cv_preds(is_induction ~ qk_frob)

thresh <- 0.5
p %>% mutate(pred = as.integer(prob >= thresh)) %>% count(is_induction, pred)  # confusion matrix

p %>% mutate(pred = as.integer(prob >= thresh)) %>%
  summarise(tp = sum(pred==1 & is_induction==1),
            fp = sum(pred==1 & is_induction==0),
            fn = sum(pred==0 & is_induction==1),
            precision = tp/(tp+fp),
            recall    = tp/(tp+fn),
            f1        = 2*precision*recall/(precision+recall))

pr_metrics <- function(formula, thresh = 0.5) {
  cv_preds(formula) %>%
    mutate(pred = as.integer(prob >= thresh)) %>%
    summarise(tp = sum(pred==1 & is_induction==1),
              fp = sum(pred==1 & is_induction==0),
              fn = sum(pred==0 & is_induction==1),
              precision = tp/(tp+fp), recall = tp/(tp+fn),
              f1 = 2*precision*recall/(precision+recall))
}

pr_metrics(is_induction ~ qk_frob)            # Frobenius
pr_metrics(is_induction ~ qk_eig)             # eigenvalue
pr_metrics(is_induction ~ qk_eig + qk_frob)   # both
```


## Over training

```{r merged_over_training}
# Per-seed max for PS and QK at each step (head-level metrics)
df_max_per_seed <- df_merged_qk_and_ps %>%
  filter(model == "pythia-14m") %>%
  group_by(model, n_params, step_modded, seed, seed_name) %>%
  summarise(
    max_ps = max(mean_ps, na.rm = TRUE),
    max_qk = max(composed_qk_score, na.rm = TRUE),
    ar_acc = first(mean_ar_acc),  # AR is model-level, same for all heads
    .groups = "drop"
  )

# Average across seeds within each model
df_avg <- df_max_per_seed %>%
  group_by(model, n_params, step_modded) %>%
  summarise(
    `Induction (PS)` = mean(max_ps),
    `Max QK` = mean(max_qk),
    `Associative Recall (Acc)` = mean(ar_acc),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(`Induction (PS)`, `Max QK`, `Associative Recall (Acc)`),
               names_to = "metric",
               values_to = "value")

ggplot(df_avg, aes(x = step_modded, y = value, color = metric)) +
  geom_line(size = 1.2) +
  geom_point(size = 1.5) +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 1.2) +
  labs(x = "Training Step (Log10)",
       y = "Score",
       color = "") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = my_colors) +
  facet_wrap(~reorder(model, n_params))



### With lines per seed
# Per-seed long format (thin lines)
df_per_seed_long <- df_max_per_seed %>%
  rename(`Induction (PS)` = max_ps,
         `Max QK` = max_qk,
         `Associative Recall (Acc)` = ar_acc) %>%
  pivot_longer(cols = c(`Induction (PS)`, `Max QK`, `Associative Recall (Acc)`),
               names_to = "metric",
               values_to = "value")

# Average across seeds (bold lines) — your existing df_avg
df_avg <- df_max_per_seed %>%
  group_by(model, n_params, step_modded) %>%
  summarise(
    `Induction (PS)` = mean(max_ps),
    `Max QK` = mean(max_qk),
    `Associative Recall (Acc)` = mean(ar_acc),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(`Induction (PS)`, `Max QK`, `Associative Recall (Acc)`),
               names_to = "metric",
               values_to = "value")

ggplot() +
  # thin per-seed lines
  geom_line(data = df_per_seed_long,
            aes(x = step_modded, y = value,
                color = metric, group = interaction(metric, seed_name),
                linetype = metric),
            alpha = 0.3, size = 0.5) +
  # bolded average lines
  geom_line(data = df_avg,
            aes(x = step_modded, y = value, color = metric, linetype = metric),
            size = 1.8) +
  geom_point(data = df_avg,
             aes(x = step_modded, y = value, color = metric, linetype = metric),
             size = 2) +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 1) +
  labs(x = "Training Step (Log10)",
       y = "Score",
       color = "",
       linetype = "") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = c("Induction (PS)" = "#0072B2",
                              "Max QK" = "#CC79A7",
                              "Associative Recall (Acc)" = "#009E73")) +
  scale_linetype_manual(values = c("Max QK" = "solid",
                                   "Induction (PS)" = "solid",
                                   "Associative Recall (Acc)" = "dotted")) +
  facet_wrap(~reorder(model, n_params))



```

## Over training (Frob)

```{r merged_over_training_frob}
# Per-seed max for PS and QK at each step (head-level metrics)
df_max_per_seed <- df_merged_qk_and_ps %>%
  filter(model == "pythia-14m") %>%
  group_by(model, n_params, step_modded, seed, seed_name) %>%
  summarise(
    max_ps = max(mean_ps, na.rm = TRUE),
    max_qk = max(composed_qk_frob, na.rm = TRUE),
    ar_acc = first(mean_ar_acc),  # AR is model-level, same for all heads
    .groups = "drop"
  )



### With lines per seed
df_per_seed_long <- df_max_per_seed %>%
  rename(`Induction (PS)` = max_ps,
         `Max QK` = max_qk,
         `Associative Recall (Acc)` = ar_acc) %>%
  pivot_longer(cols = c(`Induction (PS)`, `Max QK`, `Associative Recall (Acc)`),
               names_to = "metric",
               values_to = "value")


df_avg <- df_max_per_seed %>%
  group_by(model, n_params, step_modded) %>%
  summarise(
    `Induction (PS)` = mean(max_ps),
    `Max QK` = mean(max_qk),
    `Associative Recall (Acc)` = mean(ar_acc),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(`Induction (PS)`, `Max QK`, `Associative Recall (Acc)`),
               names_to = "metric",
               values_to = "value")

ggplot() +
  # thin per-seed lines
  geom_line(data = df_per_seed_long,
            aes(x = step_modded, y = value,
                color = metric, group = interaction(metric, seed_name),
                linetype = metric),
            alpha = 0.3, size = 0.5) +
  # bolded average lines
  geom_line(data = df_avg,
            aes(x = step_modded, y = value, color = metric, linetype = metric),
            size = 1.8) +
  geom_point(data = df_avg,
             aes(x = step_modded, y = value, color = metric, linetype = metric),
             size = 2) +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 1) +
  labs(x = "Training Step (Log10)",
       y = "Score",
       color = "",
       linetype = "") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = c("Induction (PS)" = "#0072B2",
                              "Max QK" = "#CC79A7",
                              "Associative Recall (Acc)" = "#009E73")) +
  scale_linetype_manual(values = c("Max QK" = "solid",
                                   "Induction (PS)" = "solid",
                                   "Associative Recall (Acc)" = "dotted")) +
  facet_wrap(~reorder(model, n_params))



### Rescaled
# rescale each metric to [0,1] (timing comparison, not magnitude)
scales_df <- df_per_seed_long %>%
  group_by(model, metric) %>%
  summarise(vmin = min(value, na.rm = TRUE),
            vmax = max(value, na.rm = TRUE), .groups = "drop")

rescale_metric <- function(d) {
  d %>% left_join(scales_df, by = c("model", "metric")) %>%
    mutate(value = (value - vmin) / (vmax - vmin)) %>%
    select(-vmin, -vmax)
}

df_per_seed_long_s <- rescale_metric(df_per_seed_long)
df_avg_s           <- rescale_metric(df_avg)

ggplot() +
  geom_line(data = df_per_seed_long_s,
            aes(x = step_modded, y = value,
                color = metric, group = interaction(metric, seed_name),
                linetype = metric),
            alpha = 0.3, size = 0.5) +
  geom_line(data = df_avg_s,
            aes(x = step_modded, y = value, color = metric, linetype = metric),
            size = 1.8) +
  geom_point(data = df_avg_s,
             aes(x = step_modded, y = value, color = metric, linetype = metric),
             size = 2) +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 1) +
  labs(x = "Training Step (Log10)", y = "Normalized score (per metric)",
       color = "", linetype = "") +
  theme_minimal() +
  theme(text = element_text(size = 15), legend.position = "bottom") +
  scale_color_manual(values = c("Induction (PS)" = "#0072B2",
                                "Max QK" = "#CC79A7",
                                "Associative Recall (Acc)" = "#009E73")) +
  scale_linetype_manual(values = c("Max QK" = "solid",
                                   "Induction (PS)" = "solid",
                                   "Associative Recall (Acc)" = "dotted")) +
  facet_wrap(~reorder(model, n_params))


```

## Over training (altogether)

```{r merged_all_over_training}

# both QK metrics + PS + AR per seed/step
df_max_per_seed <- df_merged_qk_and_ps %>%
  filter(model == "pythia-14m") %>%
  group_by(model, n_params, step_modded, seed, seed_name) %>%
  summarise(
    max_ps      = max(mean_ps, na.rm = TRUE),
    max_qk_eig  = max(composed_qk_score, na.rm = TRUE),   # eigenvalue metric
    max_qk_frob = max(composed_qk_frob,  na.rm = TRUE),   # Frobenius metric
    ar_acc      = first(mean_ar_acc),
    .groups = "drop"
  )

metric_order <- c("Max QK (Frob)", "Max QK (eig)",
                  "Induction (PS)", "Associative Recall (Acc)")

df_per_seed_long <- df_max_per_seed %>%
  rename(`Induction (PS)`           = max_ps,
         `Max QK (eig)`             = max_qk_eig,
         `Max QK (Frob)`            = max_qk_frob,
         `Associative Recall (Acc)` = ar_acc) %>%
  pivot_longer(cols = all_of(metric_order),
               names_to = "metric", values_to = "value") %>%
  mutate(metric = factor(metric, levels = metric_order))

df_avg <- df_per_seed_long %>%
  group_by(model, n_params, step_modded, metric) %>%
  summarise(value = mean(value), .groups = "drop")

metric_cols <- c("Max QK (Frob)" = "#CC79A7",
                 "Max QK (eig)"  = "#7B3F9E",
                 "Induction (PS)" = "#0072B2",
                 "Associative Recall (Acc)" = "#009E73")
metric_lts  <- c("Max QK (Frob)" = "solid",
                 "Max QK (eig)"  = "dashed",
                 "Induction (PS)" = "solid",
                 "Associative Recall (Acc)" = "dotted")

plot_traj <- function(per_seed, avg, ylab) {
  ggplot() +
    geom_line(data = per_seed,
              aes(step_modded, value, color = metric,
                  group = interaction(metric, seed_name), linetype = metric),
              alpha = 0.3, size = 0.5) +
    geom_line(data = avg,
              aes(step_modded, value, color = metric, linetype = metric),
              size = 1.8) +
    geom_point(data = avg,
               aes(step_modded, value, color = metric, linetype = metric),
               size = 2) +
    scale_x_log10() +
    geom_vline(xintercept = 1000, linetype = "dotted", size = 1) +
    scale_color_manual(values = metric_cols) +
    scale_linetype_manual(values = metric_lts) +
    labs(x = "Training Step (Log10)", y = ylab, color = "", linetype = "") +
    theme_minimal() +
    theme(text = element_text(size = 15), legend.position = "bottom") +
    facet_wrap(~reorder(model, n_params))
}

# --- raw ---
plot_traj(df_per_seed_long, df_avg, "Score")

# --- normalized per metric ---
scales_df <- df_per_seed_long %>%
  group_by(model, metric) %>%
  summarise(vmin = min(value, na.rm = TRUE), vmax = max(value, na.rm = TRUE),
            .groups = "drop")
rescale_metric <- function(d) {
  d %>% left_join(scales_df, by = c("model", "metric")) %>%
    mutate(value = (value - vmin) / (vmax - vmin)) %>% select(-vmin, -vmax)
}
plot_traj(rescale_metric(df_per_seed_long), rescale_metric(df_avg),
          "Normalized score (per metric)")

```


#### Tracking individual heads

```{r temporal_ordering_ind_heads}
top_induction_heads <- df_merged_qk_and_ps %>%
  filter(model == "pythia-14m", step == 143000) %>%
  group_by(seed_name) %>%
  slice_max(mean_ps, n = 1, with_ties = FALSE) %>%
  select(seed_name, layer, head) %>%
  rename(target_layer = layer, target_head = head)

top_head_trajectories <- df_merged_qk_and_ps %>%
  filter(model == "pythia-14m") %>%
  inner_join(top_induction_heads,
             by = c("seed_name",
                    "layer" = "target_layer",
                    "head" = "target_head"))

top_head_long <- top_head_trajectories %>%
  select(seed_name, step_modded, composed_qk_score, mean_ps) %>%
  rename(`QK` = composed_qk_score, `PS` = mean_ps) %>%
  pivot_longer(cols = c(QK, PS), names_to = "metric", values_to = "score")

avg_top_head <- top_head_long %>%
  group_by(step_modded, metric) %>%
  summarise(mean_score = mean(score), .groups = "drop")

ggplot() +
  geom_line(data = top_head_long,
            aes(x = step_modded, y = score,
                color = metric, group = interaction(metric, seed_name)),
            alpha = 0.3, size = 0.5) +
  geom_line(data = avg_top_head,
            aes(x = step_modded, y = mean_score, color = metric),
            size = 1.8) +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted") +
  labs(x = "Training Step (Log10)",
       y = "Score",
       color = "Metric",
       title = "Per-seed trajectory of the top induction head (identified at step 143000)") +
  theme_minimal() +
  scale_color_manual(values = c("PS" = "#0072B2",
                              "QK" = "#CC79A7")) +
  theme(text = element_text(size = 14), legend.position = "bottom")


panel_labels <- top_head_trajectories %>%
  distinct(seed_name, layer, head) %>%
  mutate(label = paste0("L", layer, "H", head - 1))  # subtract 1 if you want 0-indexed

ggplot(top_head_long, aes(x = step_modded, y = score, color = metric)) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 0.8) +
  geom_text(data = panel_labels,
            aes(x = 1, y = 0.95, label = label),
            inherit.aes = FALSE,
            hjust = 0, size = 4, color = "gray30") +
  labs(x = "Training Step (Log10)",
       y = "Score",
       color = "Metric",
       title = "Top induction head per seed: QK vs PS trajectory") +
  scale_color_manual(values = c("PS" = "#0072B2",
                              "QK" = "#CC79A7")) +
  theme_minimal() +
  theme(text = element_text(size = 13),
        legend.position = "bottom") +
  facet_wrap(~ seed_name, ncol = 3)

```

#### Tracking individual heads (Frob)

```{r temporal_ordering_ind_heads_frob}
top_induction_heads <- df_merged_qk_and_ps %>%
  filter(model == "pythia-14m", step == 143000) %>%
  group_by(seed_name) %>%
  slice_max(mean_ps, n = 1, with_ties = FALSE) %>%
  select(seed_name, layer, head) %>%
  rename(target_layer = layer, target_head = head)

top_head_trajectories <- df_merged_qk_and_ps %>%
  filter(model == "pythia-14m") %>%
  inner_join(top_induction_heads,
             by = c("seed_name",
                    "layer" = "target_layer",
                    "head" = "target_head"))

top_head_long <- top_head_trajectories %>%
  select(seed_name, step_modded, composed_qk_frob, mean_ps) %>%
  rename(`QK` = composed_qk_frob, `PS` = mean_ps) %>%
  pivot_longer(cols = c(QK, PS), names_to = "metric", values_to = "score")

avg_top_head <- top_head_long %>%
  group_by(step_modded, metric) %>%
  summarise(mean_score = mean(score), .groups = "drop")

ggplot() +
  geom_line(data = top_head_long,
            aes(x = step_modded, y = score,
                color = metric, group = interaction(metric, seed_name)),
            alpha = 0.3, size = 0.5) +
  geom_line(data = avg_top_head,
            aes(x = step_modded, y = mean_score, color = metric),
            size = 1.8) +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted") +
  labs(x = "Training Step (Log10)",
       y = "Score",
       color = "Metric",
       title = "Per-seed trajectory of the top induction head (identified at step 143000)") +
  theme_minimal() +
  scale_color_manual(values = c("PS" = "#0072B2",
                              "QK" = "#CC79A7")) +
  theme(text = element_text(size = 14), legend.position = "bottom")


panel_labels <- top_head_trajectories %>%
  distinct(seed_name, layer, head) %>%
  mutate(label = paste0("L", layer, "H", head - 1))  # subtract 1 if you want 0-indexed

ggplot(top_head_long, aes(x = step_modded, y = score, color = metric)) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 0.8) +
  geom_text(data = panel_labels,
            aes(x = 1, y = 0.95, label = label),
            inherit.aes = FALSE,
            hjust = 0, size = 4, color = "gray30") +
  labs(x = "Training Step (Log10)",
       y = "Score",
       color = "Metric",
       title = "Top induction head per seed: QK vs PS trajectory") +
  scale_color_manual(values = c("PS" = "#0072B2",
                              "QK" = "#CC79A7")) +
  theme_minimal() +
  theme(text = element_text(size = 13),
        legend.position = "bottom") +
  facet_wrap(~ seed_name, ncol = 3)

```



## Correlation over training

We measure the correlation between QK and PS at each checkpoint. We find sharp increases at step ~1000; we also see a brief dip at step ~512, perhaps because QK increases before PS.

```{r merged_cor_over_training}
# Per-seed max for PS and QK at each step (head-level metrics)
df_cor_by_seed <- df_merged_qk_and_ps %>%
  filter(model == "pythia-14m") %>%
  group_by(step_modded, model, n_params, seed_name) %>%
  summarise(r_spearman = cor(mean_ps, composed_qk_score, method = "spearman"))

summary_avg = df_cor_by_seed %>%
  group_by(model, n_params, step_modded) %>%
  summarise(
    mean_across_seeds = mean(r_spearman)
  )

grand_avg <- df_cor_by_seed %>%
  group_by(step_modded) %>%
  summarise(grand_mean = mean(r_spearman))

df_cor_by_seed %>%
  ggplot(aes(x = step_modded,
             y = r_spearman,
             color = factor(seed_name))) +
  geom_line(size = .6, alpha = .4) +   # individual curves
  geom_line(data = summary_avg, 
            aes(x = step_modded, y = mean_across_seeds), 
            color = "black", size = 2) +      # per-model average
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Correlation between QK and PS",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(9, option = "mako", 
                                                   begin = 0.8, end = 0.15))
```


### Further testing with cross-correlation

Here, we test whether QK precedes PS by asking about asymmetries in the lagged correlation. 

```{r}
# Build the per-seed time series
df_series <- df_merged_qk_and_ps %>%
  filter(model == "pythia-14m") %>%
  group_by(model, seed, seed_name, step_modded) %>%
  summarise(
    max_qk = max(composed_qk_score, na.rm = TRUE),
    max_ps = max(mean_ps, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(model, seed, step_modded)

# Lagged cross-correlation: positive lag = QK leads PS
compute_xcorr <- function(df_one_seed, max_lag = 4) {
  qk <- df_one_seed$max_qk
  ps <- df_one_seed$max_ps
  n <- length(qk)
  
  lags <- -max_lag:max_lag
  cors <- sapply(lags, function(k) {
    t_start <- max(1, 1 - k)
    t_end   <- min(n, n - k)
    if (t_end - t_start < 2) return(NA_real_)
    
    qk_window <- qk[t_start:t_end]
    ps_window <- ps[(t_start + k):(t_end + k)]
    cor(qk_window, ps_window, method = "spearman")
  })
  
  data.frame(lag = lags, correlation = cors)
}

xcorr_results <- df_series %>%
  group_by(model, seed, seed_name) %>%
  group_modify(~ compute_xcorr(.x)) %>%
  ungroup()

# Average across seeds within each model
xcorr_avg <- xcorr_results %>%
  group_by(model, lag) %>%
  summarise(mean_cor = mean(correlation, na.rm = TRUE), .groups = "drop")

# Plot
ggplot(xcorr_results, aes(x = lag, y = correlation, color = seed_name)) +
  geom_line(alpha = 0.4) +
  geom_line(data = xcorr_avg, aes(x = lag, y = mean_cor),
            color = "red", size = 1.5, inherit.aes = FALSE) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  labs(x = "Lag (positive = QK leads PS, in checkpoint units)",
       y = "Cross-correlation",
       title = "Lagged correlation between max QK and max PS") +
  theme_minimal() +
  theme(text = element_text(size = 14)) +
  facet_wrap(~model)



######### Do it again with Frob
# Build the per-seed time series
df_series <- df_merged_qk_and_ps %>%
  filter(model == "pythia-14m") %>%
  group_by(model, seed, seed_name, step_modded) %>%
  summarise(
    max_qk = max(composed_qk_frob, na.rm = TRUE),
    max_ps = max(mean_ps, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(model, seed, step_modded)

# Lagged cross-correlation: positive lag = QK leads PS
compute_xcorr <- function(df_one_seed, max_lag = 4) {
  qk <- df_one_seed$max_qk
  ps <- df_one_seed$max_ps
  n <- length(qk)
  
  lags <- -max_lag:max_lag
  cors <- sapply(lags, function(k) {
    t_start <- max(1, 1 - k)
    t_end   <- min(n, n - k)
    if (t_end - t_start < 2) return(NA_real_)
    
    qk_window <- qk[t_start:t_end]
    ps_window <- ps[(t_start + k):(t_end + k)]
    cor(qk_window, ps_window, method = "spearman")
  })
  
  data.frame(lag = lags, correlation = cors)
}

xcorr_results <- df_series %>%
  group_by(model, seed, seed_name) %>%
  group_modify(~ compute_xcorr(.x)) %>%
  ungroup()

# Average across seeds within each model
xcorr_avg <- xcorr_results %>%
  group_by(model, lag) %>%
  summarise(mean_cor = mean(correlation, na.rm = TRUE), .groups = "drop")

# Plot
ggplot(xcorr_results, aes(x = lag, y = correlation, color = seed_name)) +
  geom_line(alpha = 0.4) +
  geom_line(data = xcorr_avg, aes(x = lag, y = mean_cor),
            color = "red", size = 1.5, inherit.aes = FALSE) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  labs(x = "Lag (positive = QK leads PS, in checkpoint units)",
       y = "Cross-correlation",
       title = "Lagged correlation between max QK Frob and max PS") +
  theme_minimal() +
  theme(text = element_text(size = 14)) +
  facet_wrap(~model)
```

### Capturing onsets

```{r onsets_configurational}
combined <- df_per_seed_long %>%
  filter(metric != "Associative Recall (Acc)") %>%
  group_by(model, n_params, seed_name, metric, step_modded) %>%
  summarise(max_val = max(value, na.rm = TRUE), .groups = "drop")

df_dratio <- combined %>%
  group_by(model, seed_name, metric) %>%
  mutate(scaled = (max_val - min(max_val)) / (max(max_val) - min(max_val))) %>% 
  group_by(model, n_params, seed_name, metric) %>%
  arrange(step_modded, .by_group = TRUE) %>%
  mutate(log_step = log10(step_modded),
         d_ratio  = c(NA, diff(scaled) / diff(log_step))) %>%
  ungroup()


onset_at_deriv <- function(df, T) {
  df %>% filter(d_ratio > T) %>%
    group_by(model, n_params, seed_name, metric) %>%
    slice_min(step_modded, with_ties = FALSE) %>% ungroup() %>%
    transmute(model, n_params, seed_name, metric, onset_step = step_modded, threshold = T)
}
deriv_sweep <- map_dfr(seq(0.2, 1, by = 0.01), ~ onset_at_deriv(df_dratio, .x))



### Now average across ratios
df_onset_long = deriv_sweep %>% 
  group_by(model, n_params, seed_name, metric) %>%
  summarise(m_onset = mean(onset_step - 1)) 


df_onset_long %>%
  ggplot(aes(x = metric,
             y = m_onset)) +
  geom_jitter(size = 5, alpha = .7, width = .05) +
  scale_color_manual(values = viridisLite::viridis(4, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  scale_y_log10() +
  theme_minimal() +
  coord_flip() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  labs(x = "Metric",
       y = "Mean Deriv. Step (Log10)",
       color = "") 



mod_full = lmer(data = df_onset_long,
                 log10(m_onset) ~ metric + 
                  (1  |seed_name))
summary(mod_full)


offsets = df_onset_long %>%
  pivot_wider(id_cols = c(model, seed_name, n_params),
              values_from = m_onset,
              names_from = metric) %>%
  mutate(temporal_offset_eig = `Induction (PS)` - `Max QK (eig)`,
         temporal_offset_frob = `Induction (PS)` - `Max QK (Frob)`) %>%
  mutate(eig_before = temporal_offset_eig > 0,
         frob_before = temporal_offset_frob > 0)
mean(offsets$eig_before)
mean(offsets$frob_before)

summary(offsets$temporal_offset_eig)
summary(offsets$temporal_offset_frob)
```



# Does it also capture 1-back heads?

## Get partners

```{r}
partners_all = df_pythia_models %>%
  filter(step == 143000) %>%
  select(seed_name, layer, head, composed_qk_score, partner_layer, partner_head)


# top induction head per seed, defined by PS
partners_of_top_heads <- df_merged_qk_and_ps %>%
  filter(step == 143000) |>
  filter(model == "pythia-14m") %>%
  group_by(seed_name) |>
  slice_max(mean_ps, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(seed_name, layer, head, composed_qk_score, partner_layer, partner_head, mean_ps)

partners_of_top_heads

```

## Load 1-back data

```{r}
directory_path <- "../../data/processed/attention_summaries/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
length(csv_files)
csv_list <- csv_files %>%
  map(~ read_csv(.))


df_pythia_models <- bind_rows(csv_list) %>%
  mutate(model = sub("^EleutherAI/", "", mpath)) %>%
  filter(Layer > 0)
nrow(df_pythia_models)


### Recode as "partenr head/layer" to join with QK scores.
df_pythia_models_1back = df_pythia_models %>%
  filter(model == "pythia-14m") %>%
  mutate(partner_layer = Layer,
         partner_head = Head - 1) %>%
  select(partner_head, partner_layer, step_modded, seed_name, mean_1back)

```

## Are partner heads 1-back heads?

```{r}
# the heads that were selected as a top-induction-head's K-composition partner
partner_keys <- partners_of_top_heads %>%
  distinct(seed_name, partner_layer, partner_head) %>%
  mutate(is_top_partner = TRUE)

# every head at the final step, flagged for partner membership
df_1back_flagged <- df_pythia_models_1back %>%                      
  left_join(partner_keys) %>% #   onto this df's head identity
  mutate(is_top_partner = coalesce(is_top_partner, FALSE))


df_1back_flagged %>%
  filter(step_modded == 143001) %>%
  ggplot(aes(x = is_top_partner,
             y = mean_1back)) +
  geom_boxplot() +
  theme_minimal()


df_1back_flagged %>%
  filter(step_modded == 143001) %>%
  group_by(is_top_partner) %>%
  summarise(mean_1back_attn = mean(mean_1back),
            sd = sd(mean_1back),
            n = n())


per_seed <- df_1back_flagged %>%
  group_by(step_modded, is_top_partner, seed_name) %>%
  summarise(mean_1back_attn = mean(mean_1back), .groups = "drop")


avg <- per_seed %>%
  group_by(step_modded, is_top_partner) %>%
  summarise(mean_1back_attn = mean(mean_1back_attn), .groups = "drop")

ggplot() +
  geom_line(data = per_seed,
            aes(x = step_modded, y = mean_1back_attn,
                color = is_top_partner,
                group = interaction(is_top_partner, seed_name)),
            alpha = 0.25, size = 0.5) +
  geom_line(data = avg,
            aes(x = step_modded, y = mean_1back_attn, color = is_top_partner),
            size = 1.5) +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 1) +
  scale_color_manual(
    values = c("FALSE" = "gray60", "TRUE" = "#0072B2"),
    labels = c("FALSE" = "Other heads",
               "TRUE"  = "Partner of top\ninduction head")) +
  labs(x = "Training Step (Log10)", y = "Mean 1-back attention", color = NULL) +
  theme_minimal() +
  theme(text = element_text(size = 13), legend.position = "bottom")

final_step = df_1back_flagged %>%
  filter(step_modded == 143001) 

mod_full = lmer(data = final_step,
                mean_1back ~ is_top_partner + (1|seed_name))
summary(mod_full)

```


